Customizing Software by Direct Manipulation of Tabular Data

### 2020-05-22 paper.tex
gs
3.1

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,10pt,anonymous,review]{acmart}

@@

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we show how the behavior of a software application can
  be extended and adapted by direct manipulation, using table-driven
  customization, a new paradigm that allows end users to customize
  applications without writing traditional code.

  Instead, users directly manipulate a tabular view of the structured
  data inside the application---rather than writing imperative scripts
  (as in most customization tools). This simple model also accommodates
  a spreadsheet formula language and custom data editing widgets, which
  provide sufficient expressivity to implement many useful
  customizations.

  We illustrate the approach with Wildcard, a browser extension that
  implements table-driven customization in the context of web
  applications. Through concrete examples, we show that our paradigm can
  be used to create useful customizations for real applications. We
  share reflections from experience using the Wildcard system, on both
  its strengths and limitations relative to other customization
  approaches. Finally, we explore how this paradigm might lead to new
  software architectures that encourage this form of customization.
\end{abstract}

@@

Most end user customization systems offer a simplified version of
programming. Some scripting languages \citep{bolin2005, cook2007} have a
friendly syntax that resembles natural language. Some visual
customization tools eliminate text syntax entirely. Macro recorders
\citep{cook2007, chasins2018, anupam2000} remove some of the initial
programming burden by letting a user start with concrete demonstrations.

Despite their many differences, these approaches all share something in
common: an imperative programming model, with statement sequencing,
mutable variables and loops. End users express their ideas in scripts
made up of sequences of commands.

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the \emph{de facto} standard in GUIs today, but when it
comes to customizing software, it is rarely to be found. In this work,
we ask: what would it look like to build a software customization
interface that relies on direct manipulation? We take inspiration from
spreadsheets and visual database query interfaces
\citep{2020a, bakke2016}, which have successfully enabled end users to
run queries and computations through direct manipulation of data.

In this paper we present a technique called \emph{table-driven
customization}, which applies ideas from visual query interfaces in the
context of software customization. An application's UI is augmented with
a table view, where the user can see and manipulate the application's
internal data . Changes in the table view result in immediate
corresponding changes to the original user interface of the application,
enabling the user to customize an application with live feedback.

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:example}, we introduce the key
ideas of table-driven customization through an example scenario.

In Section~\ref{sec:architecture}, we explain the architecture of
table-driven customization. We focus on the \emph{table adapter}
abstraction, which allows many different types of underlying data to be
bidirectionally mapped to a table. We describe several types of table
adapters we've built in Wildcard, and also describe future adapters that
are supported by the general paradigm.

We have used Wildcard to build real customizations for 11 different
websites. In Section~\ref{sec:evaluation}, we present reflections from
this process, outlining the kinds of customizations we were able to
build, limitations we encountered, and reflections on the ease of
integrating scraping logic with real websites.

@@

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: End users should be able
  to customize an application by directly examining and modifying its
  data, rather than by writing imperative scripts.
\item
  \emph{Wrapping applications for customization}: Typically, tools that
  don't rely on official extension APIs resort to offering low-level
  APIs for customization. Instead, we propose a community-maintained
  library of semantic wrappers around existing applications, enabling
  end users to work with domain objects rather than low-level
  representations.
\end{itemize}

@@

To illustrate the end user experience of table-driven customization, we
consider a real example of using the Wildcard browser extension to
customize Hacker News, a tech news aggregator. Figure 1 shows
accompanying screenshots.

@@

When the user opens Hacker News in a browser equipped with the Wildcard
extension, they see a table at the bottom of the page, listing
information about each link on the site: its title, URL, number of
upvote points, etc. (Figure~\ref{fig:hacker-news}, Note A) The end user
didn't need to do any work to produce this table; Wildcard contains a
library of adapters created by programmers for extracting data from
websites.

The user's first goal might be to change the ranking of links on the
homepage. Hacker News itself uses a ranking algorithm in which the
position of an article depends not only on its point count (a measure of
popularity), but also on how long it has been on the site. If the user
hasn't been checking the site frequently, it'seasy to miss a popular
article that has fallen down the page. Sorting the page solely by number
of points would achieve a more stable ranking, with the most upvoted
articles remaining at the top of the list until they are no longer on
the home page.

To achieve this sort order, the user simply clicks on the ``points''
column header in the table. This sorts the table view by points, and the
website UI also becomes sorted in the same order
(Figure~\ref{fig:hacker-news}, Note B). This sort order persists across
page loads, so every time the user loads Hacker News, they will see
articles with this stable ordering.

Next, the user might decide to attempt a more substantial customization:
to show estimated read times next to each article, as additional context
when deciding what to read. The table contains additional columns to the
right of the original attributes for each link, where the user can enter
spreadsheet-style formulas to compute derived values. The user enters a
formula (Figure~\ref{fig:hacker-news}, Note C):

@@

This calls a built-in function that takes a URL as an argument and makes
a request to a public web service that returns an estimated read time
for the contents of that link. The result of evaluating the formula is
simply a numerical value in seconds, which is not a particularly legible
format. The user enters another formula in the next column
(Figure~\ref{fig:hacker-news}, Note D):

@@

This converts seconds to minutes by dividing by 60 and rounding to the
nearest integer, and then concatenates a label to the number. Finally,
the user chooses to show the results of this new column in the original
page (Figure~\ref{fig:hacker-news}, Note E). Each article on the page
now shows an annotation with the estimated read time in minutes.

The user might also manually add notes to the table without using
formulas. In this case, the user jots down a few notes about articles
they might want to read, and the notes appear in the page as well.

The features presented in this scenario can support a broad range of
other customizations. Section~\ref{sec:evaluation} explains many other
use cases, but first we explain how the system works internally.

@@

\textbf{Sorting/filtering}: When the user sorts or filters the query
view, an ordered list of visible IDs is sent to each table adapter. The
DOM scraping adapter uses this information to modify the list of visible
rows in the web page.

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,screen,10pt,anonymous,review]{acmart}

@@

\author{Tyler Millis}
\affiliation{%
  \institution{Massachusetts Institute of Technology}
  \city{Cambridge, MA}
  \country{USA}
}
\email{tmillis@mit.edu}

\author{Jessica Quaye}
\affiliation{%
  \institution{Massachusetts Institute of Technology}
  \city{Cambridge, MA}
  \country{USA}
}
\email{jquaye@mit.edu}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we show how the behavior of a software application can
  be extended and adapted by direct manipulation. We introduce
  \emph{table-driven customization}, a new paradigm where end users
  directly manipulate a table view of the structured data inside an
  application, rather than writing imperative scripts (as in most
  customization tools). This simple model also accommodates a
  spreadsheet formula language and custom data editing widgets, which
  provide sufficient expressivity to implement many useful
  customizations.

  We illustrate the approach with Wildcard, a browser extension that
  implements table-driven customization in the context of existing web
  applications. Through concrete examples, we show that this paradigm
  can support useful customizations for real websites. We share
  reflections from our experiences using the Wildcard system, on both
  its strengths and limitations relative to other customization
  approaches. Finally, we explore how this paradigm might lead to new
  software architectures that encourage this form of customization.
\end{abstract}

@@

Most end user customization systems offer a simplified version of
programming. Some scripting languages \citep{bolin2005, cook2007} have a
friendly syntax that resembles natural language. Other visual
customization tools eliminate text syntax entirely. Macro recorders
\citep{cook2007, chasins2018, anupam2000} remove some of the initial
programming burden by letting a user start with concrete demonstrations.
Despite their many differences, these approaches all share something in
common: an imperative programming model, with statement sequencing,
mutable variables and loops.

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the \emph{de facto} standard in GUIs today, but when it
comes to customizing those GUIs, it is rarely to be found. As a result,
switching from using software to customizing it poses a learning barrier
for users not familiar with programming, and requires an abrupt shift in
mental model even for those familiar with scripting.

In this work, we ask: what would it look like to build a direct
maniulation interface that lowers the threshold for starting to
customize software? We take inspiration from spreadsheets and visual
database query interfaces \citep{bakke2016, 2020a}, which have
successfully enabled millions of end users to compute with data through
direct manipulation.

In this paper we present a technique called \emph{table-driven
customization} inspired by those tools. An application's UI is augmented
with a table view, where the user can see and manipulate the
application's internal data. Changes in the table view result in
immediate corresponding changes to the original user interface, enabling
the user to customize an application with live feedback.

\begin{figure*}
\hypertarget{fig:overview}{%
\centering
\includegraphics[width=\textwidth]{media/overview.eps}
\caption{An overview of table-driven customization}\label{fig:overview}
}
\end{figure*}

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:example}, we illustrate the end
user experience of table-driven customization through an example
scenario using Wildcard.

In Section~\ref{sec:architecture}, we explain the system architecture of
table-driven customization. We focus on the \emph{table adapter}
abstraction, which allows many different types of underlying data to be
bidirectionally mapped to a table. We describe several types of table
adapters we've built in Wildcard, and also describe future adapters
supported by the general paradigm.

We have successfully used Wildcard to build customizations for 11
different websites. In Section~\ref{sec:evaluation}, we present
reflections from this process, outlining the kinds of customizations we
were able to build, limitations we encountered, and the challenges of
integrating scraping logic with websites.

@@

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: We agree with MacLean et
  al \citep{maclean1990} that customization should follow a ``gentle
  slope'', where users can gradually invest in more complex techniques
  to achieve deeper customizations. We believe an important point on
  this slope is the ability to customize an application by directly
  seeing and changing its internal structured data, rather than by
  writing imperative scripts.
\item
  \emph{Wrapping applications for customization}: Typically, tools that
  don't rely on official extension APIs resort to offering low-level
  APIs for customization. Instead, we propose a community-maintained
  library of semantic wrappers around existing applications, enabling
  end users to work with domain objects rather than low-level
  representations.
\end{itemize}

@@

To illustrate the end user experience of table-driven customization, we
consider an example scenario of customizing Hacker News, a tech news
aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.

@@

\textbf{Opening the table}: When the user opens Hacker News in a browser
equipped with the Wildcard extension, they see a table at the bottom of
the page. It contains a row for each link on the homepage, listing
information like the title, URL, submitter username, number of points,
and number of comments (Figure~\ref{fig:hacker-news}, Note A). The end
user didn't need to do any work to create this table, because a
programmer previously created an adapter to extract data from this
particular website, and contributed it to a shared library of adapters
integrated into Wildcard.

\textbf{Sorting by points}: First, the user decides to change the
ranking of links on the homepage. Hacker News itself uses a ranking
algorithm in which the position of an article depends not only on its
point count (a measure of popularity), but also on how long it has been
on the site. If the user hasn't been checking the site frequently, it's
easy to miss a popular article that has fallen lower on the list.
Sorting the page just by points would achieve a more stable ranking.

To achieve this ordering, the user simply clicks on the ``points''
column header in the table. This sorts the table view by points, and the
website UI also becomes sorted in the same order
(Figure~\ref{fig:hacker-news}, Note B)---Wildcard has manipulated the
website's DOM to synchronize it with the sort order of the table. This
sort order is also persisted in the browser and reapplied automatically
the next time the user loads the page, so they can always browse the
latest set of links using this sort order.

\textbf{Adding estimated read times}: Next, the user decides to attempt
a more substantial customization: to add estimated read times to
articles on Hacker News.

The table contains additional columns to the right of the original
attributes for each row, where the user can enter spreadsheet-style
formulas to compute derived values. We provide generic names for these
columns by default as spreadsheets do, so that users don't need to
choose column names themselves. The user enters a formula into the
\texttt{user1} column: (Figure~\ref{fig:hacker-news}, Note C):

@@

This calls a built-in function that takes a URL as an argument and uses
a public web API to compute an estimated read time for the link's
contents. The argument \texttt{link} refers to a column name in the
table; for each row, the formula is evaluated with the link for that
particular row. The result of evaluating this formula is a numerical
value in seconds for each article.

The user clicks the \texttt{user1} column header to sort the articles on
the page in descending order of estimated read time, helping to
prioritize reading deeper content. It would be useful to show these read
times in the page too, but before doing that, the user wants to make the
time estimates more readable. They enter another formula in the next
column, \texttt{user2}:

@@

This formula converts seconds to minutes by dividing by 60 and rounding
to the nearest integer, and then concatenates a label to the number,
producing strings like ``21 min read''.

Finally, the user clicks a menu option in the table header to display
the contents of this new column in the original page
(Figure~\ref{fig:hacker-news}, Note D). Each article on the page now
shows an annotation with the estimated read time in minutes. (The format
of how annotations appear for each item was determined by the programmer
who created the Wildcard adapter for Hacker News.)

\textbf{Adding notes to links}: The user can also manually add notes to
the table, by simply entering values into the table without using
formulas. In this case, the user jots down a few notes in another column
about articles they might want to read, and the notes appear in the page
next to the read times (Figure~\ref{fig:hacker-news}, Note D).

\textbf{Filtering out visited links}: Another way to use formulas to
customize Hacker News is to filter out articles the user has already
read. (We omit this example from the figure for brevity.) The user can
call a built-in function that returns a boolean depending on whether a
URL is in the browser's history:

\begin{verbatim}
=Visited(link)
\end{verbatim}

They can then filter the table to only contain rows where this formula
column contains \texttt{false}; visited rows are hidden both from the
table view and the original page. This is an example of a customization
that the original website could not implement, since websites don't have
access to the browser history for privacy reasons. But by using
Wildcard, the user can implement the customization locally, without
needing to expose their entire browser history to Hacker News.

This scenario has shown a few examples of how table-driven
customizations can help a user improve their experience of a website.
Section~\ref{sec:evaluation} explains many other use cases, but first we
explain how the system works internally.

@@

\textbf{Sorting/filtering}: When the user sorts or filters the query
view, an ordered list of visible IDs is sent to each table adapter. The
DOM scraping adapter uses this information to change the list of rows
shown in the web page.
